---
title: "workshop"
author: "Quentin Le Roux"
date: "11/17/2020"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
library(dplyr)
library(ggplot2)
library(reshape)
library(gapminder)
library(purrr)
```

# Exercise 1

1. Download the dataset and show the summary.

```{r dataset_1, echo=FALSE}
data <- read.csv("winequality-red.csv", sep=";", header=TRUE)
class(data)
summary(data)
```

2. Add the variable “label” for alcohol percentage (alcohol variable) to group the alcohol concentration in the following way:

Light < 9; 9 ≤ Medium < 12 ; Strong ≥ 12

```{r labels}
data$label = ""
data$label[data$alcohol<9]="Light"
data$label[9<=data$alcohol & data$alcohol<12]="Medium"
data$label[data$alcohol>=12]="Strong"
```

3. How many Light, Medium and Strong wines there are in the dataset?

```{r label_count, echo=FALSE}
table(data$label)
```

4. Create a barplot for the 3 levels variable “label” where the color of bars depends on the level of the alcohol concentration, with title: “Alcohol concentration”

```{r barplot_1, echo=FALSE}
ggplot(data, aes(x=as.factor(label), fill=as.factor(label) )) +  
  geom_bar(alpha=0.6) +
  scale_fill_brewer(palette = "Set1", name="Alcohol concentration") +
  xlab("Alcohol level") + ylab("Count") + 
  ggtitle("Alcohol concentration")
```

5. Show histograms for fixed.acidity by label

```{r hist_1, echo=FALSE}
ggplot(data, aes(x=fixed.acidity, fill=label)) +
  geom_histogram(alpha=0.6, binwidth=1) + #, position = 'identity'
  scale_fill_brewer(palette = "Set1", name="Alcohol concentration") +
  labs(fill="") + xlab("Acidity") + ylab("Frequency")
```

6. Create a density plot taking into account the citric.acid for the three level of the variable “label”

```{r density_1, echo=FALSE}
ggplot(data, aes(x=citric.acid, fill=label)) +
  geom_density(alpha=0.2) +
  scale_fill_brewer(palette = "Set1") +
  xlab("Citric Acid Amount") + ylab("Density") + 
  ggtitle("Citric Acid Amount Between Kind of Red Wines") + 
  theme_classic()
```

7. Create a boxplot for the overall dataset (Consider the Log Scale)

```{r boxplot_1, echo=FALSE}
logData <- data %>% map_if(is.numeric, function(x) log1p(x)) %>%
  as.data.frame
meltData <- melt(logData)
ggplot(meltData, aes(factor(variable), value)) +
  geom_boxplot() + facet_wrap(~variable, scale="free") + 
  xlab("Variables") + ylab("Values (Log Scale)") +
  ggtitle("Boxplot Overview of the Red Wine Dataset")

```

8. Create a scatter plot quality and alcohol where the color of points varies depending on the following quality ranges (2,4],(4,6],(6,8]. Add a dashed line for the median. Give a title to the plot and to the x and y axes (Use geom_jitter() instead of geom_point())

```{r scatter_1, echo=FALSE}
new_data <- data %>% mutate(quality_range=case_when(
  quality>2 & quality<=4 ~ "(2,4]",
  quality>4 & quality<=6 ~ "(4,6]",
  quality>6 ~ "(6,8]"
))
ggplot(new_data, aes(quality, alcohol)) + geom_jitter(aes(colour = quality_range)) + 
  xlab("Quality Range") + ylab("Alcohol Content") +
  ggtitle("Scatter Plot Overview of the Quality of Red Whine depending on Alcohol") + stat_summary(fun.y=median, geom="line", colour="black", linetype=2)
```


# Exercise 2

1. Load the dataset gapminder from the package gapminder

```{r dataset_2, echo=FALSE}
gapdata <- gapminder
```

2. Create a tibble 5×4 containing the number of observation per continent, the number of countries per continent and the average of life expectation.

```{r tibble, echo=FALSE}
tibbled_data <- gapdata %>% group_by(continent) %>% summarize(
  observations = n(),
  number_countries = n_distinct(country),
  average_lifespan = mean(lifeExp)
)
tibbled_data
```

3. Show the columns of country, year, lifeExp and add a variable that represents the year of life expectancy gained with respect to 1957, for each country, until 1963.

```{r lifespan_gain, echo=FALSE}
lifespan_gain <- gapdata %>% select(country, year, lifeExp) %>%
  filter(year == 1957 | year == 1962) %>% #no 1963 date
  group_by(country) %>%
  mutate(lifeExp_gain_since_1957 = lifeExp - sum(if_else(year==1957,lifeExp,0))) %>%
  filter(year != 1957)
lifespan_gain
```

4. What was the most populous European country in 1992? Modify the population variable by dividing it by 106. Show the results sorted according to population in a descending order.

The most populous country was :
```{r pop_max, echo=FALSE}
pop_max <- gapdata  %>% filter(year==1992) %>% arrange(desc(pop))
head(pop_max, 1)[1]
```

```{r pop_max_table, echo=FALSE}
pop_max <- pop_max %>% mutate(pop=pop/106)
pop_max
```

5. Draw a scatterplot of gdpPercap and lifeExp where the color of points depends on the continent. Overall also a smooth line (loess) with its confidence intervals. Give the plot a title and axis labels. *Hint: log scale x axis*

```{r scatter_2, echo=FALSE}
logPopMax <- pop_max %>% map_at("gdpPercap", function(x) log1p(x)) %>%
  as.data.frame
ggplot(logPopMax, aes(gdpPercap, lifeExp)) + geom_point(aes(colour = continent)) + geom_smooth(method = loess) + 
  xlab("GDP per Capita (Log Scale)") + ylab("Life Expectancy") +
  ggtitle("Life Expectancy vs. GDP per Capital in 1992")
```

# Exercise 3
1. Select a dataset and variables to be used for clustering:
  - Iris dataset, variable Species not included.
  - mtcars dataset, considering only the following variables : ‘mpg’,‘disp’, ‘wt’.

```{r dataset_3, echo=FALSE}
iris_data <- iris %>% select(-"Species")
mtcars_data <- mtcars %>% select("mpg", "disp", "wt")
available_datasets <- c("iris", "mtcars")
```

2. Indicate the desired number of clusters to be used and display the results in a scatterplot where colors depend on clusters. *Hint : add runtime: shiny to the YAML-Header.*

```{r shinapp, echo=FALSE}
shinyApp(

# Define UI for application that draws a histogram
ui <- fluidPage(
    
    pageWithSidebar(
        headerPanel('K-means clustering example'),
        sidebarPanel(
          uiOutput("selIn1"),
          uiOutput("selIn2"),
          uiOutput("selIn3"),
          uiOutput("clusters")
        ),
        mainPanel(
          plotOutput('plot1')
        )
    )
    
),

# Define server logic required to draw a histogram
server <- function(input, output) {
    
    output$selIn1=renderUI({
      selectInput('dataset', 'Select a dataset', available_datasets,
                  selected=available_datasets[1])
    })
    
    output$selIn2 <- renderUI({
      if (input$dataset=="iris"){
        dat = iris_data
      } else {
        dat = mtcars_data
      }
      selectInput('xcol', 'Select first variable for clustering',
                  colnames(dat), selected=colnames(dat)[1])
    })
    
    output$selIn3 <- renderUI({
      if (input$dataset=="iris"){
        dat = iris_data
      } else {
        dat = mtcars_data
      }
      selectInput('ycol', 'Select second variable for clustering',
                  colnames(dat), selected=colnames(dat)[2])
    })
    
    output$clusters <- renderUI({
        numericInput('clusters', 'Number of clusters', 
                     3, min = 1, max = 9)
    })
    
    selectedData <- reactive({
      if (input$dataset=="iris"){
        dat = iris_data
      } else {
        dat = mtcars_data
      }
      dat[, c(input$xcol, input$ycol)]
    })
    
    clusters <- reactive({
        kmeans(selectedData(), input$clusters)
    })
    
    output$plot1 <- renderPlot({
        palette(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3",
                  "#FF7F00", "#FFFF33", "#A65628", "#F781BF", "#999999"))
        par(mar=c(5.1, 4.1, 4.1, 8.1), xpd=TRUE)
        plot(selectedData(),
             col = clusters()$cluster,
             pch = 17, cex = 2, alpha = 0.5)
        points(clusters()$centers, col="black", pch = 17, cex = 3, lwd = 5)
        legend("topright", title="Clusters", inset=c(-0.3, 0), legend=c(paste("Cluster", 1:input$clusters),"Centroids"), pch=c(rep.int(17,input$clusters),17), col=c(1:input$clusters,"black"))
    })
    
},
  
  options = list(height = 800)
)
```