---
title: "lecture_4_statistical_inference_2"
author: "QLR"
date: "11/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Exercise 3

Let $X_1,...,X_n\sim t_3$ where $n=25$. Let $\theta=T(F)=\frac{(q_{.75}-q_{.25})}{1.34}$ where $q_p$ denotes the $p^{th}$ quantile. Do a simulation to compare the coverage and length of the following confidence intervals for $\theta$: 

- Normal interval with standard error from the bootstrap
- Boostrap percentile interval
- pivotal bootstrap interval

## Simulating our sample confidence interval

### Declaring the estimator function:

```{r estimator_function}
estimator <- function(dat) {
  est <- (quantile(dat, 0.75)-quantile(dat, 0.25))/1.34
  return(est)
}
```

### Simulating our random variables:

```{r simulation}
# Estimating Theta
n <- 25
data <- rt(n, 3)
theta_hat <- estimator(data)

# declaring a confidence interval
alpha <- 0.05
```

### Bootstrapping from our sample

```{r bootstraping}
B=1000
theta_record <- vector(length=B)

for (i in 1:B) {
  dat_ <- sample(data, replace=TRUE)
  theta_record[i] <- estimator(dat_)
}
```

an approximate $1-\alpha$ confidence interval for $T(F)$ is $T(\hat{F}_n)\pm \mathcal{z}_{\alpha/2}.\hat{se}$

### Normal-based interval

For a 95 percent confidence interval $\mathcal{z}_{\alpha/2}=\mathcal{z}_{.05/2}\approx 1.96$

```{r normal_based}
se = sd(theta_record)
z = qnorm(1-alpha/2)
confidence_interval_norm = c(theta_hat - z*se, theta_hat + z*se)

confidence_interval_norm
```

### Bootstrap percentile confidence interval

The bootstrap percentile interval is defined by $C_n=(\theta^\ast_{\alpha/2}, \theta^\ast_{1-\alpha/2})$

```{r bootstrap_percentile}
bootstrap_percentile_interval = c(quantile(theta_record, alpha/2), 
                                  quantile(theta_record, 1-alpha/2))

bootstrap_percentile_interval
```

### Bootstrap pivotal confidence interval

The bootstrap percentile interval is defined by $C_n=(2\hat{\theta}_n-\hat{\theta}^\ast_{1-\alpha/2}, 2\hat{\theta}_n-\hat{\theta}^\ast_{\alpha/2})$

```{r bootstrap_pivotal}
bootstrap_pivotal_interval = c(2*theta_hat - quantile(theta_record, 1-alpha/2),
                               2*theta_hat - quantile(theta_record, alpha/2))

bootstrap_pivotal_interval
```

## Comparing the coverage

We do a series of simulation to compare the coverage and length of the previous confidence interval

```{r comparing_coverage}
n_2=100
theta_hat_2=c()
for (i in 1:1000){
  data_2 = rt(n_2, 3)
  theta_hat_2[i] <- as.numeric(estimator(data_2))
}

check_interval <- function(x, ci) {
  N_rep <- length(x)
  ratio <- sum(x>ci[1] & x<ci[2])/N_rep
  return(ratio)
}
```

```{r comparing_nci}
check_interval(theta_hat_2, confidence_interval_norm)
```

```{r comparing_perc}
check_interval(theta_hat_2, bootstrap_percentile_interval)
```

```{r comparing_pivci}
check_interval(theta_hat_2, bootstrap_pivotal_interval)
```